A weakly structured stem for human origins in Africa

### 2023-03-13 paper.tex
final edits?

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron~P.~Ragsdale}
\author[2]{Timothy~D.~Weaver}
\author[3]{Elizabeth~G.~Atkinson}
\author[4]{Eileen~Hoal}
\author[4]{Marlo~M\"{o}ller}
\author[2,5,$\dag$,*]{Brenna~M.~Henn}
\author[6,$\dag$,**]{Simon~Gravel}
\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, Houston, TX, USA}
\affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[5]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
\affil[6]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
\affil[$\dag$]{Co-Corresponding Authors}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\date{\normalsize \today}
\maketitle

@@

    Despite broad agreement that \emph{Homo sapiens} originated within Africa,
    considerable uncertainty surrounds specific models of divergence and
    migration across the continent\citep{Henn2018-rf}. Progress is hampered by
    limited fossil and genomic data, as well as variability in prior divergence
    time estimates\citep{Henn2018-rf}.  To discriminate among such models, we
    consider linkage disequilibrium and diversity-based statistics, optimized
    for rapid, complex demographic inference \citep{Ragsdale2019-nt}.  We infer
    detailed demographic models for populations across Africa, including
    representatives from eastern and western groups and newly sequenced whole
    genomes from 44 Nama (Khoe-San) individuals. We infer a reticulated African
    population history in which contemporary population structure dates back to
    Marine Isotope Stage (MIS) 5. The earliest population divergence among
    contemporary populations occurs 120--135ka and was preceded by structure
    between two or more weakly differentiated ancestral \emph{Homo} populations
    connected by gene flow over hundreds of thousands of years. Such
    \emph{weakly structured stem} models better explain patterns of
    polymorphism that were previously attributed to contributions from archaic
    hominins in Africa
    \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
    Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to models with archaic
    introgression, we predict fossil remains from coexisting ancestral
    populations should be genetically and morphologically similar, and that an
    inferred 1--4\% of genetic differentiation among contemporary human
    populations can be attributed to genetic drift between stem populations. We
    show that model misspecification explains variation in previous divergence
    time estimates and argue that studying a suite of models is key to robust
    inferences about deep history.

@@

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent studies
have shown that introducing an archaic hominin ``ghost'' population contributing to 
African populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models, mostly in western Africa
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
Lorente-Galdos2019-vz,Durvasula2020-td},
%Plagnol: YRI 
%Here we present a new method for addressing whether archaic human groups contributed to the modern gene pool (called ancient admixture),
%we find strong evidence for ancient admixture in both a European and a West African population
%
%Hammer: Mandenka, Biaka, and San. Title:
%"Genetic evidence for archaic admixture in Africa"
%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
%
%Hey: Strongest signal in Yoruba, Baka
% Abstract: "consistent with human/archaic divergence."
%
%Ragsdale: YRI
%abstract: We estimate that an unidentified, deeply diverged population admixed with modern humans within Africa both before and after the split of African and Eurasian populations, contributing 4 − 8% genetic ancestry to individuals in world-wide populations.
%
%Lorente Galdos: Khoesan, Mbuti, W. AFR
%title:"reveals archaic gene flow"
%
%Durvasula: West African
%title: "Recovering signals of ghost archaic introgression"
%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans." 
but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} 
and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}
populations.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence\citep{Hsieh2016-gk}. However, these prior
studies share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped to clarify early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

@@

We quantify migration rates among populations after their divergence
$\sim120$ka. Prior to agripastoralist expansion $5$ka, migration between the
ancestors of the Nama and other groups is an order of magnitude weaker than
that observed between western and eastern Africans (Table
\ref{tab:migration-rates}). All models infer relatively high gene flow between
eastern and western Africa ($m\approx2\times10^{-4}$, the constant proportion
of immigrant lineages per generation over $60$ka). We further find that
Back-to-Africa gene flow at the beginning of the Holocene primarily affected
the ancestors of the Ethiopian agricultural populations
\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
estimated to be approximately $65\%$. We observe significant gene flow from the
Amhara and Oromo into the Nama, a signal which is likely a proxy for the
movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a $25\%$ ancestry
contribution 2,000 years ago. While this gene flow is not apparent from the
\texttt{ADMIXTURE} plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
into the Khoe-San component which has drifted appreciably from its ancestral
eastern African source. Colonial period admixture from Europeans into the Nama
was estimated at $15\%$, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\sim120$ka. The second event
$\sim100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of eastern/western Africans including the ancestors of
``Out-of-Africans''. Reticulated models do not have a unique and 
well-defined ``basal'' human population divergence.
We suggest conceptualizing the events $120$ka as the time of most recent
shared ancestry among sampled populations.
However, interpreting population divergence times in population genetics is 
always challenging due
to the co-estimation of divergence time and subsequent migration;
methods assuming clean vs reticulated splits can infer different split dates
(Figure~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of divergence time in the
literature\citep{Bergstrom2021-iw,Henn2018-rf}.

@@

We also find that Stem 2 continued to contribute to western Africans during the
Last Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an
interpretation is reinforced by the differential migration rates between
regions, i.e., the gene flow from Stem 2 to western Africans is estimated to be
five times that of the rate to eastern Africans during this period. We
performed a variety of validation tests to explore sensitivity of our
assumptions, including relaxing fixed parameters (SI
Section~\ref{sec:validation_strategy}). Most validation tests resulted in
parameters similar to the models discussed above. However, one exception was
the inferred Out-of-Africa and eastern/western African divergences which were
10--15ka younger than our fixed parameters. These younger dates are clearly at 
odds with the accepted timing of Out-of-Africa expansion at approximately $\>50$ka, 
as based on archaeological, climatic and paleo-morphological information
\citep{Fu2014-am,Groucutt2015-do,Prufer2021-ao,Beyer2021-ym}. 
We interpret the free estimate for eastern African vs. European divergence as reflecting 
our inclusion of only a single Out-of-Africa population in the model, the lack of a nearby 
source for back-to-Africa gene flow, and other regionally complex parameters. 
However, older pan-African features of our inferred models are weakly affected by 
this analysis choice (SI Section~\ref{sec:models}).

@@

Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
archaic hominin admixture having also occurred in Africa. Previous work that sought to 
explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
archaic hominin admixture as an alternative model, by referring to additional (or ``ghost'') 
branches required to fit the data as ``archaic'' 
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,
Durvasula2020-td} and assuming (or inferring) deep divergences. These perspectives have 
oriented interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil 
(e.g., evolution of early \emph{Homo sapiens} \cite{Grun2020-su}) data. Here, we have shown 
that a model of a weakly structured stem better captures the apparently inconsistent patterns 
of polymorphisms.

Preferring weakly-structured-stem models over archaic-admixture models potentially has a 
range of implications. First, with a weakly structured stem, there is no need to posit that 
an archaic hominin population in Africa stayed reproductively isolated from the ancestral 
human lineage for hundreds of thousands of years before the initiation of gene flow. Instead, 
there would simply have been continuous or recurrent contact between two or more groups 
present in Africa.

@@

Third, multiple studies have shown a correspondence between phenotypic differentiation, 
usually assessed with measurements of the cranium, and genetic differentiation among 
human populations and between humans and Neanderthals \citep{Relethford1994-mh,
Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI Section~\ref{sec:morpho_diff}). 
This correspondence potentially allows predictions of our model to be related to the 
fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
\citep{Harvati2011-xs}, Ishango, 
Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and Nazlet Khater, 
Egypt (35-40ka) \citep{Crevecoeur2012-we}, 
have morphological features that may reflect relatively recent gene flow from archaic hominins \citep{Harvati2011-xs, Crevecoeur2016-cl}, which has been used in support of previously inferred archaic admixture events in human ancestry \cite{Scerri2018-nl, Durvasula2020-td, Bergstrom2021-iw}. 
The weakly structured stem model is not incompatible with archaic admixture having occurred in the ancestry of these fossils, but would imply, by contrast, that such individuals are unlikely to have contributed much ancestry to  contemporary humans. 
The fossil record of Africa is 
sparse during the earlier time period of the stems \aprcomment{($\gtrsim120$ka)},
but of the fossils that date to this time period, 
some are fairly similar overall in morphology to contemporary humans (e.g., Omo 1 from 
Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological 
features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in morphology to 
have been assigned to species other than \emph{Homo sapiens} (e.g., DH1 from Dinaledi, 
South Africa \citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
differences between the stems were comparable to those among contemporary human populations, 
the most morphologically divergent fossils are also unlikely to represent branches that 
contributed appreciably to contemporary human ancestries.

@@

We are grateful for the DNA contribution from each participant that enabled
this study; in particular we wish to highlight the generous participation of
the Richtersveld Nama community in South Africa and help from local research
assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
Uren and Cedric Werely. We thank the African Genome Diversity Project for data
generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
thank Luke Anderson-Trocm\'e and Georgette Femerling for assistance in creating
the map in Figure~\ref{fig:diversity} and Figure \ref{fig:toymodels},
respectively. We thank Nuria Melisa Morales-Garcia for extensive data
visualization discussion and designing Figures \ref{fig:proposed-models} and
\ref{fig:best-fit-models}. This research was supported by CIHR project grant
437576, Natural Sciences and Engineering Research Council of Canada (NSERC)
grant RGPIN-2017-04816, the Canada Research Chair program to SG, and the Canada
Foundation for Innovation. This research was supported by an NIH grant
R35GM133531 to BMH. The content is solely the responsibility of the authors and
does not necessarily represent the official views of the National Institutes of
Health. MM and EH acknowledge the support of the DSI-NRF Centre of Excellence
for Biomedical Tuberculosis Research, South African Medical Research Council
Centre for Tuberculosis Research, Division of Molecular Biology and Human
Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University,
Cape Town, South Africa.

@@

\subsection*{Gene genalogy reconstruction}

@@

\begin{figure}[ht!]
    \centering
    %\includegraphics{figures/samples-diversity.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:diversity}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    %\includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see SI Section~S\ref{sec:f4}
        for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last $100$ka.
    }
    \label{fig:predictions}
\end{figure}

>>>>

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron~P.~Ragsdale}
\author[2]{Timothy~D.~Weaver}
\author[3]{Elizabeth~G.~Atkinson}
\author[4]{Eileen~Hoal}
\author[4]{Marlo~M\"{o}ller}
\author[2,5,$\dag$,*]{Brenna~M.~Henn}
\author[6,$\dag$,**]{Simon~Gravel}
\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
\affil[2]{Department of Anthropology, University of California, Davis, CA, USA}
\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, Houston, TX, USA}
\affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[5]{UC Davis Genome Center, University of California, Davis, CA, USA}
\affil[6]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
\affil[$\dag$]{Co-Corresponding Authors}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\date{\normalsize \today}
\maketitle

@@

    Despite broad agreement that \emph{Homo sapiens} originated within Africa,
    considerable uncertainty surrounds specific models of divergence and
    migration across the continent\citep{Henn2018-rf}. Progress is hampered by
    limited fossil and genomic data, as well as variability in prior divergence
    time estimates\citep{Henn2018-rf}.  To discriminate among such models, we
    consider linkage disequilibrium and diversity-based statistics, optimized
    for rapid, complex demographic inference \citep{Ragsdale2019-nt}.  We infer
    detailed demographic models for populations across Africa, including
    representatives from eastern and western groups and newly sequenced whole
    genomes from 44 Nama (Khoe-San) individuals. We infer a reticulated African
    population history in which contemporary population structure dates back to
    Marine Isotope Stage (MIS) 5. The earliest population divergence among
    contemporary populations occurs 120--135ka and was preceded by structure
    between two or more weakly differentiated ancestral \emph{Homo} populations
    connected by gene flow over hundreds of thousands of years. Such
    \emph{weakly structured stem} models better explain patterns of
    polymorphism that were previously attributed to contributions from archaic
    hominins in Africa \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,
    Ragsdale2019-nt, Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to
    models with archaic introgression, we predict fossil remains from
    coexisting ancestral populations should be genetically and morphologically
    similar, and that an inferred 1--4\% of genetic differentiation among
    contemporary human populations can be attributed to genetic drift between
    stem populations. We show that model misspecification explains variation in
    previous divergence time estimates and argue that studying a suite of
    models is key to robust inferences about deep history.

@@

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.  However, new
population genetic tools now allow for inference involving tens to hundreds of
genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent studies have shown
that introducing an archaic hominin ``ghost'' population contributing to
African populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models, mostly in western Africa
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
Lorente-Galdos2019-vz,Durvasula2020-td},
%Plagnol: YRI Here we present a new method for addressing whether archaic human
%groups contributed to the modern gene pool (called ancient admixture), we find
%strong evidence for ancient admixture in both a European and a West African
%population
%
%Hammer: Mandenka, Biaka, and San. Title: "Genetic evidence for archaic
%admixture in Africa" Abstract:"allow us to infer that contemporary African
%populations contain a small proportion of genetic material (≈2%) that
%introgressed ≈35 kya from an archaic population that split from the ancestors
%of anatomically modern humans ≈700 kya."
%
%Hey: Strongest signal in Yoruba, Baka Abstract: "consistent with human/archaic
%divergence."
%
%Ragsdale: YRI abstract: We estimate that an unidentified, deeply diverged
%population admixed with modern humans within Africa both before and after the
%split of African and Eurasian populations, contributing 4 − 8% genetic
%ancestry to individuals in world-wide populations.
%
%Lorente Galdos: Khoesan, Mbuti, W. AFR title:"reveals archaic gene flow"
%
%Durvasula: West African title: "Recovering signals of ghost archaic
%introgression" Abstract:"Our analyses of site frequency spectra indicate that
%these populations derive 2 to 19% of their genetic ancestry from an archaic
%population that diverged before the split of Neanderthals and modern humans." 
but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} and central
African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}
populations.  This has driven speculation about the geographic range of this
ghost population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence\citep{Hsieh2016-gk}. However, these prior studies
share two weaknesses. First, they only contrast a single-origin model with an
archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).  Second, they focus on
a small subset of African diversity, either because of small sample sizes (2-5
genomes) or because they rely on 1000 Genomes data which only recruited
populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C).  While ancient DNA from Eurasia has helped to
clarify early human history outside of Africa, there is no comparably ancient
DNA to elucidate early history in Africa \citep{Lipson2022-xf}.

@@

We quantify migration rates among populations after their divergence
$\sim120$ka. Prior to agripastoralist expansion $5$ka, migration between the
ancestors of the Nama and other groups is an order of magnitude weaker than
that observed between western and eastern Africans (Table
\ref{tab:migration-rates}). All models infer relatively high gene flow between
eastern and western Africa ($m\approx2\times10^{-4}$, the constant proportion
of immigrant lineages per generation over $60$ka). We further find that
back-to-Africa gene flow at the beginning of the Holocene primarily affected
the ancestors of the Ethiopian agricultural populations
\citep{Molinaro2019-rf}, comprising close to 65\% of their genetic ancestry. We
observe significant gene flow from the Amhara and Oromo into the Nama, a signal
which is likely a proxy for the movement of eastern African caprid and cattle
pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a
$25\%$ ancestry contribution 2,000 years ago. While this gene flow is not
apparent from the \texttt{ADMIXTURE} plot (Figure~\ref{fig:diversity}), the
ancestry is likely grouped into the Khoe-San component which has drifted
appreciably from its ancestral eastern African source. Colonial period
admixture from Europeans into the Nama was estimated at $15\%$, similar to
proportions suggested by \texttt{ADMIXTURE} (Figure~\ref{fig:diversity}).

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate ancestry
contributions from different stems in contemporary groups. We observed two
merger events. The first, between Stem 1S and Stem 2, resulted in the formation
of an ancestral Khoe-San population $\sim120$ka. The second event $\sim100$ka
between Stem 1E and Stem 2, resulted in the formation of the ancestors of
eastern/western Africans including the ancestors of ``Out-of-Africans''.
Reticulated models do not have a unique and well-defined ``basal'' human
population divergence.  We suggest conceptualizing the events at $120$ka as the
time of most recent shared ancestry among sampled populations. However,
interpreting population divergence times in population genetics is always
challenging due to the co-estimation of divergence time and subsequent
migration; methods assuming clean vs reticulated splits can infer different
split dates (Figure~\ref{fig:supp-rccr-data} and
Figure~\ref{fig:supp-IM-misspecification}). Therefore, wide variation exists in
estimates of divergence time in the
literature\citep{Bergstrom2021-iw,Henn2018-rf}.

@@

We also find that Stem 2 continued to contribute to western Africans during the
Last Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an
interpretation is reinforced by the differential migration rates between
regions, i.e., the gene flow from Stem 2 to western Africans is estimated to be
five times that of the rate to eastern Africans during this period. We
performed a variety of validation tests to explore sensitivity of our
assumptions, including relaxing fixed parameters (SI
Section~\ref{sec:validation_strategy}). Most validation tests resulted in
parameters similar to the models discussed above. However, one exception was
the inferred Out-of-Africa and eastern/western African divergences which were
10--15ka younger than our fixed parameters. These younger dates are at odds
with the accepted timing of the Out-of-Africa expansion that contributed to
later human populations at approximately $\>50$ka, as based on archaeological,
climatic and fossil information \citep{Fu2014-am,Groucutt2015-do,
Prufer2021-ao,Beyer2021-ym}. We interpret the free estimate for eastern African
vs. European divergence as reflecting our inclusion of only a single
Out-of-Africa population in the model, the lack of a nearby source for
back-to-Africa gene flow, and other regionally complex parameters. However,
older pan-African features of our inferred models are minimally affected by
this analysis choice (SI Section~\ref{sec:models}).

@@

Evidence for archaic hominin admixture in Eurasia has bolstered the
plausibility of archaic hominin admixture having also occurred in Africa.
Previous work that sought to explain patterns of polymorphism inconsistent with
a single-origin model thus focused on archaic hominin admixture as an
alternative model, by referring to additional (or ``ghost'') branches required
to fit the data as ``archaic'' \citep{Plagnol2006-lt,Hammer2011-bx,
Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz, Durvasula2020-td} and
assuming (or inferring) deep divergences. These perspectives have oriented
interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil
(e.g., evolution of early \emph{Homo sapiens} \cite{Grun2020-su}) data. Here,
we have shown that a model of a weakly structured stem better captures the
apparently inconsistent patterns of polymorphisms.

Preferring weakly-structured-stem models over archaic-admixture models has a
range of implications. First, with a weakly structured stem, there is no need
to posit that an archaic hominin population in Africa stayed reproductively
isolated from the ancestral human lineage for hundreds of thousands of years
before the initiation of gene flow. Instead, there would simply have been
continuous or recurrent contact between two or more groups present in Africa.

@@

Third, multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh, Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also
SI Section~\ref{sec:morpho_diff}).  This correspondence potentially allows
predictions of our model to be related to the fossil record. Some \emph{Homo
sapiens} fossils, from Iho Eleru, Nigeria (13ka) \citep{Harvati2011-xs},
Ishango, Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and
Nazlet Khater, Egypt (35-40ka) \citep{Crevecoeur2012-we}, have morphological
features that may reflect relatively recent gene flow from archaic hominins
\citep{Harvati2011-xs, Crevecoeur2016-cl}, which has been used in support of
previously inferred archaic admixture events in human ancestry
\cite{Scerri2018-nl, Durvasula2020-td, Bergstrom2021-iw}.  The weakly
structured stem model is not incompatible with archaic admixture having
occurred in the ancestry of these fossils, but would imply, by contrast, that
such individuals are unlikely to have contributed much ancestry to contemporary
humans. The fossil record of Africa is sparse during the earlier time period of
the stems ($\gtrsim200$ka), but of the fossils that date to this time period,
some are fairly similar overall in morphology to contemporary humans (e.g., Omo
1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are
similar in some morphological features to contemporary humans (e.g., Irhoud 1
from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others
are different enough in morphology to have been assigned to species other than
\emph{Homo sapiens} (e.g., DH1 from Dinaledi, South Africa
\citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic
differences between the stems were comparable to those among contemporary human
populations, the most morphologically divergent fossils are also unlikely to
represent branches that contributed appreciably to contemporary human
ancestries.

@@

We are grateful for the DNA contribution from each participant that enabled
this study; in particular we wish to highlight the generous participation of
the Richtersveld Nama community in South Africa and help from local research
assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
Uren and Cedric Werely. We thank the African Genome Diversity Project for data
generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
thank Luke Anderson-Trocm\'e and Georgette Femerling for assistance in creating
the map in Figure~\ref{fig:diversity} and Figure~\ref{fig:toymodels},
respectively. We thank Nuria Melisa Morales-Garcia for extensive data
visualization discussion and designing Figures~\ref{fig:proposed-models} and
\ref{fig:best-fit-models}. This research was supported by CIHR project grant
437576, Natural Sciences and Engineering Research Council of Canada (NSERC)
grant RGPIN-2017-04816, the Canada Research Chair program to SG, and the Canada
Foundation for Innovation. This research was supported by an NIH grant
R35GM133531 to BMH. The content is solely the responsibility of the authors and
does not necessarily represent the official views of the National Institutes of
Health. MM and EH acknowledge the support of the DSI-NRF Centre of Excellence
for Biomedical Tuberculosis Research, South African Medical Research Council
Centre for Tuberculosis Research, Division of Molecular Biology and Human
Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University,
Cape Town, South Africa.

@@

\subsection*{Gene genealogy reconstruction}

@@

\begin{figure}[ht!]
    \centering
    %\includegraphics{figures/samples-diversity.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:diversity}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    %\includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is
        mostly recent.} From the best fit models of our two parameterizations
        (A and B: continuous migration, C and D: merger with stem migration),
        we predicted differentiation and shared drift between populations at
        past times. (A, C) We computed expected pairwise differences $H_{i,j}$
        between individuals sampled from populations $i$ and $j$ existing at
        time $t$. (B, D) To understand how drift between stems explains
        contemporary structure, we computed the proportion $\alpha^2$ of drift
        between pairs of sampled contemporary populations that aligns with
        drift between past populations (here Nama and Mende, see SI
        Section~S\ref{sec:f4} for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.  Most present-day differentiation
        dates back to the last $100$ka.
    }
    \label{fig:predictions}
\end{figure}

\begin{figure}[t!]
    \centering
    %\includegraphics{figures/validation}
    \caption{
        \textbf{Model validation using independent statistics.} (A--C) Using
        our best fit models to linkage disequilibrium and pairwise diversity statistics,
        we simulated expected conditional site-frequency-spectra (cSFS) and
        compared to the observed cSFS from the data. Our inferred models
        provide a good fit to the data, even though these were not used in our
        inference. Across the three populations, ancestral state
        misidentification was consistently inferred to be $1.5-1.7\%$ for
        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
        which we used to estimate coalescence rate trajectories
        and cross-coalescence rates between pairs of populations. While
        coalescence rate distributions are informative statistics about past
        evolutionary processes, interpretation can be hindered by migration and
        population structure, and translating relative cross-coalescence rate
        curves (RCCR) into population divergence times is especially prone
        to misinterpretation.
        For example, in our model, the Mende-Gumuz split occurs prior to the
        Gumuz-British split. However, the model also predicts recent elevated
        Mende-Gumuz RCCR. This pattern, also observed in the data, does not
        indicate that the Mende and Gumuz split more recently.
    }
    \label{fig:validation}
\end{figure}
